---
title: "Introduction to multiomics R/Bioconductor package"
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
bibliography: bibliography.bib
date: "`r doc_date()`"
package: "`r pkg_ver('multiomics')`"
vignette: >
  %\VignetteIndexEntry{Introduction to multiomics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```

# Introduction

In recent years, many different approaches to integrate multiomics data has
been proposed. For instance, [@Mangnier2025] benchmarked different approaches
to integrate metagenomics and metabolomics data, and evaluated different
methods based on robustness and interpretability.

In the publication, they proposed the following methods to address the following
research questions:

```{r table-methods, echo=FALSE}
library(knitr)

methods_table <- data.frame(
    `Scientific question` = c(
        "Is there any relationship between microorganisms and metabolites at a global level?",
        "Are microbiome and metabolome datasets summarizable through a limited number of components?",
        "Can we identify associations between metabolites and species?",
        "Can we identify core microorganisms and metabolites?"
    ),
    `Research aim` = c(
        "Global associations",
        "Data summarization",
        "Individual associations",
        "Feature selection (univariate)"
    ),
    `Recommended method` = c(
        "Mantel test",
        "RDA",
        "MiRKAT",
        "CODA-LASSO (compositional covariates)"
    ),
    check.names = FALSE
)

kable(methods_table, caption = "Overview of multiomics statistical approaches")
```

The goal of the `multiomics` R/Bioconductor package is to implement these
methods and more within the `MultiAssayExperiment` ecosystem, making them easily
available in analysis workflows.

# Workflow

```{r "start", message=FALSE}
library("multiomics")
```

## Installation

To install `multiomics`, install `BiocManager` first, if it is not installed.
Afterwards use the `install` function from `BiocManager`.

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("multiomics")
```

## Load `multiomics`

```{r load-packages, message=FALSE, warning=FALSE}
library("multiomics")
```

We use and example dataset from `mia` package.

```{r load-data}
library(mia)
data("HintikkaXOData")
mae <- HintikkaXOData
```

The dataset is provided as a `MultiAssayExperiment` (`MAE`) object, a container
designed for managing multi-omics data. Individual omics datasets are stored as
`(Tree)SummarizedExperiment` (`TreeSE`) objects, each representing a single
omics layer. A key feature of the `MAE` object is that samples are linked
across assays: a sample in one omics layer is associated with the
corresponding sample in all other layers. This mapping enables automatic
bookkeeping and ensures that samples remain correctly aligned during downstream
analyses.

Next we replace missing values with 0s in metabolomics layer and agglomerate
microbiome data into Family level.

```{r wrangle_data}
# Replace missing values with 0
assay(mae[[2]])[is.na(assay(mae[[2]]))] <- 0

# Agglomerate microbiota data to Family level
mae[[1]] <- agglomerateByRank(mae[[1]], rank = "Family")
```

We apply CLR transformation to microbiome data, and metabolites data is
log-transformed.

```{r transformation}
# Apply CLR and log10
mae[[1]] <- transformAssay(mae[[1]], assay.type = "counts", method = "rclr")
mae[[2]] <- transformAssay(mae[[2]], assay.type = "nmr", method = "log10", pseudocount = 0.5)
```


## Is there any relationship between microorganisms and metabolites at a global level?

To answer this question, we apply the Mantel test to CLR-transformed microbiome
data and log-transformed metabolite data. This allows us to use Euclidean
distances.

```{r mantel}
mantel <- getMantel(
    mae,
    experiments = c(1, 2),
    assay.types = c("rclr", "log10"),
    dist.methods = c("euclidean", "euclidean")
)
mantel
```

Based on the results, there is a significant association between microbiome and
metabolite profiles.


## Are microbiome and metabolome datasets summarizable through a limited number of components?

This research question can be addressed by applying redundancy analysis (RDA).
Note that this method is based on linear relationships, which requires using
CLR-transformed data.

```{r rda}
rda <- getRDA(
    mae,
    experiments = c(1, 2),
    assay.types = c("rclr", "log10")
)
rda
```

The results of RDA are usually visualized with a biplot.

```{r show_rda}
library(ggvegan)
p <- ordiggplot(rda) +
    geom_ordi_axis() +
    geom_ordi_point("sites") +
    geom_ordi_arrow("species")  +
    geom_ordi_arrow("biplot")

# Add coloring
gb <- ggplot_build(p)

sites <- dplyr::left_join(
    gb$data[[3]],
    colData(mae) |> as.data.frame(),
    by = c("label" = "Sample")
)

p + geom_point(
        data = sites,
        aes(x, y, colour = Fat),
        size = 3
    )
```

Based on the results, 96\% of the microbiome variance can be explained by
metabolites. More than 80\% of the variance is captured by the first components.

## Can we identify associations between metabolites and species?

Next, we continue with MiRKAT. The idea is to test the association of each
metabolite with the microbiome profile.

```{r mirkat}
mirkat <- getMiRKAT(
    mae,
    experiments = c(1, 2),
    assay.types = c("rclr", "log10"),
    dist.methods = c("euclidean", "manhattan", "canberra")
)
mirkat |> head()
```

We can visualize the p-values for instance with a bar plot.

```{r visualize_mirkat}
ggplot(mirkat, aes(x = reorder(rownames(mirkat), desc(omnibus_p)), y = omnibus_p)) +
    geom_col() +
    coord_flip() +
    theme_minimal() +
    geom_hline(yintercept = 0.05, color = "red", linewidth = 1) +
    labs(x = NULL, y = "omnibus p-value")
```

For instance, _rownames(mirkat)[ order(mirkat[["omnibus_p_adj"]]) ][[1L]]_ is
strongly associated with the microbiome.

## Can we identify core microorganisms and metabolites?

We can use LASSO to identify core taxa and metabolites by compressing the data.
To follow the approach in [@Calle2023], we apply the method to log-ratio
transformed data.

```{r coda_lasso}
coda_lasso <- getLASSO(
    mae,
    experiments = c(1, 2),
    assay.types = c("counts", "log10"),
    apply.logratio = TRUE
)
```

To visualize these relationships, we use a heatmap of the coefficient matrix.
Positive and negative coefficients indicate positive and negative associations.

```{r show_coda_lasso}
# Put into long format for ggplot
library(tidyr)
df_long <- coda_lasso |>
    as.data.frame() |>
    tibble::rownames_to_column("Taxa") |>
    pivot_longer(
        cols = -Taxa,
        names_to = "Metabolite",
        values_to = "Coefficient"
    )

library(ggplot2)
ggplot(df_long, aes(x = Metabolite, y = Taxa, fill = Coefficient)) +
    geom_tile() +
    scale_fill_gradient2(
        low = "blue",
        mid = "white",
        high = "red",
        midpoint = 0
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Are there shared latent drivers in microbiome and metabolite data?

Robust Joint RPCA identifies shared latent structure across datasets such as
microbiome and metabolite profiles. It first learns a low-rank joint
representation that captures shared sample-level factors while allowing
dataset-specific loadings, reducing noise and improving robustness. PCA is then
applied to the reconstructed representation to extract the main shared
components for interpretation.

```{r robust_joint_pca}
# Apply data transformations. With impute=FALSE, missing values are preserved
# and not imputed.
mae[[1]] <- transformAssay(
    mae[[1]], assay.type = "counts", method = "rclr", impute = FALSE,
    name = "rclr_wo_imputation")

# Run joint-RPCA
res <- getJointRPCA(
   mae,
   experiments = c(1, 2),
   assay.types = c("rclr_wo_imputation", "log10")
)
```

The result can be visualize with a scatter plot similarly to regular PCA.

```{r visualize_joint_rpca}
res <- cbind(res, colData(mae)[ , "Fat", drop = FALSE])

library(ggplot2)

res <- res |> as.data.frame()

ggplot(res, aes(x = PC1, y = PC2, color = Fat)) +
    geom_point(size = 3, alpha = 0.9) +
    theme_minimal() +
    labs(x = "PC1", y = "PC2", color = "Fat")
```

# Citing `multiomics`

We hope that `r Biocpkg("multiomics")` will be useful for your research. Please
use the following information to cite the package and the overall approach.
Thank you!

```{r citation}
## Citation info
citation("multiomics")
```

# `R` session information.

```{r session_info}
sessionInfo()
```
